---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# epitraxr

<!-- badges: start -->

[![ForeSITE Group](https://github.com/EpiForeSITE/software/raw/e82ed88f75e0fe5c0a1a3b38c2b94509f122019c/docs/assets/foresite-software-badge.svg)](https://github.com/EpiForeSITE) [![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental) [![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://github.com/EpiForeSITE/epitraxr/blob/master/LICENSE.md) [![R-CMD-check](https://github.com/EpiForeSITE/epitraxr/actions/workflows/r-check.yaml/badge.svg)](https://github.com/EpiForeSITE/epitraxr/actions/workflows/r-check.yaml) [![pkgdown](https://github.com/EpiForeSITE/epitraxr/actions/workflows/pkgdown.yaml/badge.svg)](https://github.com/EpiForeSITE/epitraxr/actions/workflows/pkgdown.yaml) [![codecov](https://codecov.io/gh/EpiForeSITE/epitraxr/graph/badge.svg?token=KC48SFPX39)](https://codecov.io/gh/EpiForeSITE/epitraxr)

<!-- badges: end -->

## Overview

EpiTrax is a central repository for epidemiological data developed by Utah State's Department of Health and Human Services (DHHS). It is now used by several other states. Through EpiTrax, public health officials have access to many different types of disease surveillance data, which they use to produce regular (e.g., weekly, monthly, annual) reports on their respective jurisdictions. This can be a tedious, time-intensive process, often involving multiple spreadsheets.

The `epitraxr` package makes it fast and easy to process EpiTrax data and generate reports. With `epitraxr` it is simple to setup a report pipeline, then you simply hit "run" and select your latest EpiTrax export---`epitraxr` will do the rest!

## Installation

You can install the development version of `epitraxr` from [GitHub](https://github.com/EpiForeSITE/epitraxr) with:

``` r
# install.packages("devtools")
devtools::install_github("EpiForeSITE/epitraxr")
```

For local health departments, we recommend going through the steps under [Installation For Local Health Departments](#installation-for-local-health-departments) (below). Our experience is that LHD-managed machines need additional tools and packages installed first. This list will walk you through those steps.

### Installing Package Dependencies

Many features of `epitraxr` will work with the standard installation. However, there are several optional features, such as PDF report generation and the R Shiny app, that require you to install additional dependencies. You can do that with the command:

``` r
# install.packages("devtools")
devtools::install_github("EpiForeSITE/epitraxr", dependencies = TRUE)
```

For PDF reports, you may need to install a LaTeX distribution as well. We recommend using [TinyTeX](https://yihui.org/tinytex/) as described in the [R Markdown Cookbook](https://bookdown.org/yihui/rmarkdown-cookbook/install-latex.html). First, install the `tinytex` package:

``` r
install.packages("tinytex")
```

Then, install TinyTeX (the LaTeX distribution) using the `tinytex` package:

``` r
tinytex::install_tinytex()
# to uninstall TinyTeX, run
# tinytex::uninstall_tinytex()
```

After that, you can use `epitraxr` to generate PDF reports.

### Installation For Local Health Departments {#installation-for-local-health-departments}

In our experience, machines managed by local health departments (LHDs) are less likely to have installed all necessary packages and tools. For these users, we recommend completing the following steps before installing the package using the instructions above.

1.  Install the [latest version of R](https://cran.rstudio.com)
2.  [Install RStudio](https://posit.co/download/rstudio-desktop/)
3.  If on Windows, install the version of [Rtools](https://cran.r-project.org/bin/windows/Rtools/) corresponding to the version of R you installed. This is needed to build and install certain types of packages.
4.  Install `devtools` (needed to install the package from GitHub)

``` r
install.packages("devtools")
```

5.  Install the `tinytex` package

``` r
install.packages("tinytex")
```

6.  Install TinyTeX (the LaTeX distribution) using the `tinytex` package

``` r
tinytex::install_tinytex()
```

7.  Install `epitraxr`

-   **Note:** If you previously attempted to install `epitraxr` and ran into an error (such as one solved by the above steps), you will need to install the package again with the option `force = TRUE`. This ensures the full package is downloaded and installed properly.

``` r
devtools::install_github("EpiForeSITE/epitraxr", dependencies = TRUE, force = TRUE)
```

-   This command will also update the package (e.g., if you want to get the latest version from GitHub.).

## Usage

`epitraxr` can be used in either "standard" mode or "piped" mode (recommended).

### Standard Mode

```{r usage-standard}
library(epitraxr)

data_file <- system.file(
  "sample_data/sample_epitrax_data.csv",
  package = "epitraxr"
)
epitrax_data <- read_epitrax_data(data_file)

report <- create_report_annual_counts(
  data = epitrax_data,
  disease_names = c("Chickenpox", "Measles", "Lyme disease")
)

report
```

### Piped Mode (recommended)

```{r usage-piped}
library(epitraxr)

config_file <- system.file(
  "sample_data/sample_config.yml",
  package = "epitraxr"
)

disease_list_file <- system.file(
  "sample_data/sample_disease_list.csv",
  package = "epitraxr"
)

epitrax <- get_epitrax(data_file) |>
  epitrax_set_config_from_file(config_file) |>
  epitrax_add_report_diseases(list(
    internal = disease_list_file,
    public = disease_list_file
  )) |>
  epitrax_ireport_annual_counts()

epitrax$internal_reports$annual_counts
```

Piped mode makes it super simple to add additional reports.

```{r usage-piped-expanded}
epitrax <- get_epitrax(data_file) |>
  epitrax_set_config_from_file(config_file) |>
  epitrax_add_report_diseases(list(
    internal = disease_list_file,
    public = disease_list_file
  )) |>
  epitrax_ireport_annual_counts() |>
  epitrax_ireport_monthly_avgs() |>
  epitrax_ireport_ytd_counts_for_month()

list(epitrax$internal_reports)
```

### Shiny App

The epitraxr package includes a Shiny app, which provides a user-friendly interface for running report generation operations with epitraxr.

To run the Shiny app, you'll need to install the optional dependencies, such as `shiny` and `DT`.

``` r
devtools::install_github("EpiForeSITE/epitraxr", dependencies = TRUE)
```

Next, load the package.

``` r
library(epitraxr)
```

Then you can run the app with `run_app()`.

``` r
run_app()
```

### Scripts

We've developed a couple of handy scripts that are included in the [GitHub repo](https://github.com/EpiForeSITE/epitraxr) under the [`scripts/` folder](https://github.com/EpiForeSITE/epitraxr/tree/main/scripts):

-   [`epitraxr-piped.R`](https://github.com/EpiForeSITE/epitraxr/blob/main/scripts/epitraxr-piped.R): Showcases report generation using epitraxr's Piped Mode **(recommended)**
-   [`epitraxr-generator.R`](https://github.com/EpiForeSITE/epitraxr/blob/main/scripts/epitraxr-generator.R): Showcases report generation using the epitraxr's Standard Mode

These scripts provide you with a full report generation process that should run successfully if you have the package installed with all additional dependencies (see above). They provide identical output to the [original project](https://github.com/EpiForeSITE/epitrax-report-automation) (currently inactive).

### Troubleshooting: Common Usage Problems

-   If `epitraxr` unexpectedly fails to generate reports after previously succeeding, check that none of the generated reports are open on your machine. Often, an `epitraxr` script will start by deleting existing reports from folders before generating new ones. If the files are opened, `epitraxr` cannot delete them. Close any open files and try running the script again.

## Getting Help

If you encounter a clear bug, please file an issue with a minimal reproducible example on [GitHub](https://github.com/EpiForeSITE/epitraxr/issues).

## Code of Conduct

Please note that the epitraxr project is released with a [Contributor Code of Conduct](https://epiforesite.github.io/epitraxr/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.

## Acknowledgments

This project was made possible by cooperative agreement CDC-RFA-FT-23-0069 from the CDC’s Center for Forecasting and Outbreak Analytics. Its contents are solely the responsibility of the authors and do not necessarily represent the official views of the Centers for Disease Control and Prevention.
